% WildfireIQ Kamloops -- Project Report
% Compiles with pdflatex (standard packages only).
\documentclass[11pt]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{array}
\usepackage{titlesec}
\usepackage{enumitem}
\usepackage{tikz}
\usepackage{fancyhdr}
\usepackage[hidelinks]{hyperref}

\usetikzlibrary{positioning, arrows.meta}

% ==== Colour scheme (professional slate + ember accent)
\definecolor{ink}{HTML}{1F2A37}      % headings
\definecolor{ember}{HTML}{C2410C}    % accent
\definecolor{slate}{HTML}{475569}    % secondary text
\definecolor{mist}{HTML}{EEF2F6}     % light fills
\definecolor{line}{HTML}{CBD5E1}     % rules

\hypersetup{
  colorlinks=true,
  linkcolor=ember,
  urlcolor=ember,
}

% ==== Section styling
\titleformat{\section}{\color{ink}\Large\bfseries}{\thesection}{0.6em}{}
\titleformat{\subsection}{\color{ink}\large\bfseries}{\thesubsection}{0.6em}{}
\titlespacing*{\section}{0pt}{1.4em}{0.6em}

% ==== Header / footer
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\headrule}{\hbox to\headwidth{\color{line}\leaders\hrule height \headrulewidth\hfill}}
\fancyhead[L]{\small\color{slate}WildfireIQ Kamloops}
\fancyhead[R]{\small\color{slate}Project Report}
\fancyfoot[C]{\small\color{slate}\thepage}

% ==== Table helpers
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\renewcommand{\arraystretch}{1.25}

% ==== A simple coloured callout box
\newcommand{\callout}[1]{%
  \par\medskip\noindent
  \colorbox{mist}{\parbox{\dimexpr\linewidth-2\fboxsep}{\color{ink}\small #1}}%
  \par\medskip}

\begin{document}

% ============================ COVER ============================
\begin{titlepage}
\thispagestyle{empty}
\color{ink}
\vspace*{1.2cm}

{\noindent\color{ember}\rule{\linewidth}{2.5pt}}
\vspace{0.8cm}

\noindent{\Huge\bfseries WildfireIQ Kamloops}\\[0.5cm]
{\noindent\LARGE A wildfire risk, air quality, and community preparedness platform for the Thompson-Okanagan}\\[0.6cm]

{\noindent\large\color{slate} Project Report and Final Implementation}

\vspace{0.6cm}
{\noindent\color{ember}\rule{\linewidth}{1pt}}

\vspace{2.2cm}
\noindent
\begin{tabular}{@{}ll@{}}
{\bfseries Author} & Deeparsh Singh Dang \\[0.3em]
{\bfseries Funding} & TRU Sustainability Research Grant, 2025--2026 \\[0.3em]
{\bfseries Region} & Thompson-Okanagan, British Columbia \\[0.3em]
{\bfseries Licence} & MIT (code), CC-BY-4.0 (written content) \\
\end{tabular}

\vfill
\callout{\textbf{Informational only.} This platform supports public awareness and planning. It does not replace official guidance from the BC Wildfire Service, BC Emergency Management Climate Readiness, or Environment and Climate Change Canada.}

\noindent{\color{slate}\small Generated for project handoff.}
\end{titlepage}

% ============================ TOC ============================
\newpage
\tableofcontents
\newpage

% ============================ 1. OVERVIEW ============================
\section{Project Overview}

WildfireIQ Kamloops is a web platform that brings wildfire information into one
place for people who live in the Thompson-Okanagan region of British Columbia.
It answers four everyday questions: where are the fires right now, is the air
safe to breathe, how do I get my home and family ready, and how has fire risk
changed over time.

The platform is built around a three-dimensional globe of British Columbia.
Live data layers sit on top of the globe, and three supporting pages cover air
quality, household preparedness, and long-term climate trends. Two machine
learning models add a forward-looking view: one estimates daily wildfire risk,
the other forecasts air quality for the next two days.

Everything runs on free and open data sources, stores no personal information,
and is released as open source. The goal was a tool that is genuinely useful to
a resident, a journalist, or a local official, and that is honest about what it
can and cannot tell you.

\callout{\textbf{At a glance.} 4 connected modules, 17 automated data jobs, 32
API endpoints, 2 validated machine learning models, and 23 years of historical
fire and weather records, all served locally at no cloud cost.}

% ============================ 2. WHAT IT DOES ============================
\section{What the Platform Does}

The platform has four parts. Each one is described below in plain terms.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}L{3.1cm}L{10.5cm}@{}}
\toprule
\textbf{Module} & \textbf{What it shows} \\
\midrule
\textbf{3D Globe} &
The home screen. Shows active fires, satellite heat detections, evacuation
zones, fire-weather stations, a smoke forecast you can step through hour by
hour, and an AI risk map. Layers can be turned on and off, and each one has a
plain-language explanation. \\
\textbf{Air Quality} &
A live air quality reading for Kamloops, a forecast for the next 48 hours with
an honest uncertainty range, a six-pollutant breakdown, a one-year smoke
calendar, and health guidance from Health Canada. \\
\textbf{Preparedness Hub} &
A personalised FireSmart checklist that adapts to your home type and season, a
live check of whether your neighbourhood is under an evacuation order, progress
badges, and a private share link. Nothing leaves your device. \\
\textbf{Climate Trends} &
A scrolling story of how the region's fire seasons have changed over 27 years,
with clear trend lines, and a look ahead under different emission scenarios. \\
\bottomrule
\end{tabular}
\caption{The four modules of WildfireIQ Kamloops.}
\end{table}

\subsection{The Wildfire Risk Map}

The home screen is a three-dimensional globe of British Columbia that opens
with a short camera fly-in to the Kamloops region. Six data layers can be
switched on and off from a side rail: active fires, satellite heat detections,
evacuation zones, fire-weather stations, the smoke forecast, and the AI risk
grid. Clicking any item opens a detail panel with its full record and a plain
explanation of what the layer means. The fire and evacuation lists are sorted
newest first, and the evacuation panel has a control to hide zones that have
already been called off. A search box and four camera presets let a user jump
to a place or to the wider region. The smoke layer includes a time slider that
steps through about three days of forecast, one hour at a time.

\subsection{The Air Quality Monitor}

This page leads with a large dial showing the current Air Quality Health Index
for Kamloops, colour-coded on Health Canada's scale. Below it, a chart shows the
48-hour forecast as a shaded band: the middle line is the best estimate, and the
band widens when the forecast is less certain. The page also breaks down six
pollutants, shows a province-wide map of nearby monitoring stations, and
displays a calendar of the last 365 days where each day is coloured by its worst
air quality. Health guidance is grouped into tabs for the general public,
at-risk groups, and outdoor workers, and a user can opt in to a browser
notification when air quality is forecast to cross a threshold they choose.

\subsection{The Community Preparedness Hub}

A short three-step setup asks for a neighbourhood, the type of home, and
notification preferences. From that, the hub builds a FireSmart checklist drawn
from the official Home Ignition Zone actions, filtered to the user's situation
and re-ordered so the most seasonally relevant actions come first. Completing
actions earns points and unlocks badges, and a user can attach a photo to any
action as a personal record. A live widget checks whether the chosen
neighbourhood sits inside an active evacuation zone. Everything is stored in the
browser only. A share link encodes progress in the web address itself, so it can
be shared without any data passing through a server.

\subsection{The Climate Trends page}

This page tells the region's fire-season story as the reader scrolls. It opens
with 27 years of total area burned, marks the landmark seasons, then shows clear
trend lines for summer temperature, dryness, and fire-weather severity. Later
sections look ahead under three emission scenarios and estimate how the number
of extreme fire-weather days per year could change by decade. Every chart lists
its data source and method and offers a download of the underlying numbers, and
the page prints cleanly to a short PDF for use as a reference.

\subsection{Key numbers}

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}L{6.5cm}L{6.5cm}@{}}
\toprule
\textbf{Item} & \textbf{Figure} \\
\midrule
Historical fire records (1999 to today) & 15,996 incidents \\
Daily weather history & 10,010 days (about 27 years) \\
Air quality stations shown (province-wide) & 34 \\
Fire-weather stations & 18 \\
AI risk map cells (Thompson-Okanagan) & 185 hexagons \\
FireSmart checklist actions & 30 \\
Smoke forecast steps & 73 hourly snapshots \\
Automated tests passing & 47 backend, 22 frontend \\
\bottomrule
\end{tabular}
\caption{Selected figures from the final build.}
\end{table}

% ============================ 3. DATA ============================
\section{Where the Data Comes From}

Every layer is built from a free, public data source. The platform downloads,
cleans, and stores the data, then serves it to the browser. No source requires
payment, and only three need a free sign-up key (Cesium for the globe, NASA for
satellite heat detections, and WAQI for pollutants).

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}L{3.6cm}L{5.6cm}L{3.6cm}@{}}
\toprule
\textbf{Layer} & \textbf{Source} & \textbf{Coverage} \\
\midrule
Active fires & BC Wildfire Service (DataBC) & All of BC \\
Satellite heat detections & NASA FIRMS (VIIRS, MODIS) & All of BC \\
Evacuation zones & BC Emergency Management & All of BC \\
Fire Weather Index & Open-Meteo, computed in-house & 18 BC stations \\
Air quality (live) & Environment Canada GeoMet & All of BC \\
Air quality (history) & Open-Meteo CAMS & Kamloops \\
Smoke forecast & Environment Canada RAQDPS-FW & All of BC \\
Weather & Open-Meteo & Kamloops \\
Climate projections & ClimateData.ca structure & Regional \\
Globe terrain and imagery & Cesium Ion & Global \\
\bottomrule
\end{tabular}
\caption{Data sources and the area each layer covers.}
\end{table}

\subsection{How the data stays current}

Seventeen small jobs run on timers inside the application. Active fires refresh
every 15 minutes, evacuation zones every 5 minutes, weather and air quality
every hour, and the smoke forecast every six hours. When the app starts, it
checks each source and refreshes anything older than 30 minutes, so the screen
shows current information within seconds of opening rather than waiting for the
next scheduled update.

\callout{\textbf{Honest about freshness.} A satellite heat detection is a
``worth investigating'' signal, not a confirmed fire. A high Fire Weather Index
describes dangerous conditions, not an active fire. The platform spells this out
on every layer so readers are not misled.}

\subsection{From raw data to the screen}

Each source is handled by a small job that follows the same steps: download the
latest data, drop anything malformed or out of range, reshape it into a common
format, and save it as a compressed table on disk. The browser never talks to an
outside source directly. It only reads these cleaned tables through the
platform's own interface, which keeps the app fast and means a brief outage at a
source does not break the screen.

Two pieces of data are calculated by the platform rather than downloaded. The
Fire Weather Index is computed from daily weather using Canada's official
equations, because the national service that normally publishes it was
unreachable for much of the build. The historical fire-and-climate summary, used
by the Climate Trends page, is rebuilt each night by joining 27 years of fire
records with matching weather and a full Fire Weather Index run, producing one
row of metrics per year.

A daily weather record stretching back to 1999 underpins both the risk model and
the climate page. Because the long-term archive lags real time by a few days,
the platform stitches the most recent observed days onto the end of it, so the
record always reaches today and the risk map reflects current conditions.

% ============================ 4. AI ============================
\section{How the AI Works}

Two machine learning models add a forward-looking view. Both were trained on
real historical records and tested on years the models had never seen, which is
what makes their accuracy figures trustworthy rather than memorised.

\subsection{Wildfire risk model}

This model estimates the chance that a fire starts somewhere in the region on a
given day, using the day's weather and fire-weather conditions. The 2022 and
2023 fire seasons were both kept out of training, exactly as the proposal
committed: 2022 was used for tuning and 2023 for the final test. The model
learned from the 1999 to 2021 records.

The model is a gradient-boosted decision tree (LightGBM). Each day it sees is
described by about 40 numbers: that day's temperature, humidity, wind, rainfall,
and dryness; the six Fire Weather Index codes; short and month-long trends in
those values; how long it has been since meaningful rain; and the time of year.
The model's raw output is calibrated so that a stated chance of, say, seventy
percent really does correspond to fires occurring on about seven in ten such
days, which matters for the colours shown on the map.

On the unseen 2023 test year it scored a precision-recall area of 0.66, clearly
ahead of the traditional fire-weather-threshold method at 0.52. A precision-recall
score rewards a model for ranking the genuine fire days above the quiet ones, so
a higher number means the model is better at telling busy days from calm ones.
The single regional estimate is then scaled for each map hexagon by how often
that specific area has burned in the past, which is what gives the map its local
texture. The official fire-danger class is shown beside it for comparison.

\subsection{Air quality forecaster}

This model predicts fine-particle pollution (the harmful part of smoke) for the
next 48 hours. It is built from 21 small models working together: for each of
seven look-ahead times, three of them learn the low, middle, and high end of the
likely range. Showing all three turns a single guess into an honest band that
widens when the model is less certain, rather than a confident line that could
be wrong. Its inputs are recent pollution readings, the weather, and the time of
day at Kamloops.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}L{4cm}r r@{}}
\toprule
\textbf{Forecast horizon} & \textbf{Model error} & \textbf{Simple baseline} \\
\midrule
6 hours  & 2.27 & 2.85 \\
12 hours & 3.20 & 4.06 \\
24 hours & 3.82 & 3.63 \\
48 hours & 3.32 & 3.92 \\
\bottomrule
\end{tabular}
\caption{Air quality forecast error in micrograms per cubic metre (lower is
better). The model beats the ``tomorrow equals today'' baseline at most
horizons.}
\end{table}

% ============================ 5. CLIMATE FINDINGS ============================
\section{What the Climate Record Shows}

Beyond serving live data, the platform produces a small piece of original
analysis. Using 27 years of records for the Thompson-Okanagan, it measures how
several fire-season conditions have changed. The trends are calculated with a
method that is not thrown off by a single extreme year, and each one is tested
to see whether the change is large enough to be unlikely by chance.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}L{6.2cm}L{4.0cm}L{2.8cm}@{}}
\toprule
\textbf{Measure (1999 to today)} & \textbf{Change over the period} & \textbf{Clear trend?} \\
\midrule
Average July daytime temperature & about +4.7 degrees C & Yes \\
Summer dryness of the air & rising & Yes \\
Peak fire-weather severity & rising & Yes \\
Days per year of extreme fire weather & about +50 days & Yes \\
Summer rainfall & no clear change & No \\
Length of the fire season & no clear change & No \\
\bottomrule
\end{tabular}
\caption{Measured trends for the Thompson-Okanagan. A ``clear trend'' means the
change is unlikely to be chance.}
\end{table}

The pattern is consistent and worth stating plainly: summers are getting hotter
and the air drier, and the number of days with dangerous fire weather has grown
sharply, even though total rainfall and the calendar length of the season have
not changed in a measurable way. In other words, the season is not obviously
longer, but it is more intense. The platform is careful not to overstate this:
where a trend is not statistically clear, the page says so rather than implying
a pattern that the data does not support. The future-looking sections are
labelled as estimates and use a placeholder climate dataset that has the correct
shape but is not the final download.

% ============================ 6. HOW IT IS BUILT ============================
\section{How It Is Built}

The platform has two halves: a browser front end that draws the globe and pages,
and a back end that collects data and runs the models. They talk over a simple
web interface.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  node distance=0.7cm and 1.1cm,
  box/.style={rectangle, rounded corners, draw=line, fill=mist, text=ink,
    align=center, minimum height=1.05cm, minimum width=3.7cm, font=\small},
  accent/.style={rectangle, rounded corners, draw=ember, line width=1pt,
    fill=white, text=ink, align=center, minimum height=1.05cm,
    minimum width=3.7cm, font=\small},
  arr/.style={-{Stealth[length=2.2mm]}, draw=slate, thick}
]
\node[box] (browser) {Browser\\\footnotesize React, Cesium globe, charts};
\node[accent, below=of browser] (api) {Back end\\\footnotesize FastAPI, scheduler, models};
\node[box, below left=of api] (data) {Data store\\\footnotesize Parquet, DuckDB, SQLite};
\node[box, below right=of api] (sources) {Public sources\\\footnotesize BCWS, NASA, ECCC, Open-Meteo};

\draw[arr] (browser) -- node[right, font=\scriptsize\color{slate}] {requests} (api);
\draw[arr] (api) -- (data);
\draw[arr] (sources) -- node[right, font=\scriptsize\color{slate}] {ingest} (api);
\draw[arr] (data) -- (api);
\end{tikzpicture}
\caption{High-level architecture. Public sources feed the back end on timers;
the browser reads cleaned data through the API.}
\end{figure}

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}L{3cm}L{10.5cm}@{}}
\toprule
\textbf{Layer} & \textbf{Main tools} \\
\midrule
Front end & React, TypeScript, Vite, CesiumJS (3D globe), Tailwind, Visx (charts) \\
Back end & Python, FastAPI, APScheduler (timers), LightGBM (models) \\
Storage & Parquet files, DuckDB (analytics), SQLite (run log) \\
\bottomrule
\end{tabular}
\caption{Technology choices, kept deliberately simple and free to run.}
\end{table}

The whole system runs on one machine at no cloud cost. There is no separate
database server or task queue, which keeps it easy to run and hand off. The
front end loads the globe first and fetches the other pages only when needed, so
the initial page stays small (about 69 kilobytes compressed).

\subsection{Keeping it fast and current}

A few choices keep the platform responsive. The three secondary pages are loaded
on demand rather than up front, so the first screen is light. Responses that do
not change often, such as the climate history, are marked so the browser can
reuse them instead of asking again. The analytics engine is opened once when the
app starts so the first request is quick. Most importantly, when the app starts
it refreshes any data source older than 30 minutes, which means a person opening
the platform after it has been idle still sees current fires, current air
quality, and a risk map built on today's weather, without waiting.

\subsection{Testing and quality}

The platform ships with an automated test suite: 47 checks on the back end and
22 on the front end. These confirm that each data source produces the expected
shape, that values stay within sensible bounds (for example, no negative
pollution readings and no fires placed outside the province), that the Fire
Weather Index calculation behaves correctly, and that the trend statistics
recover a known answer. The front-end checks cover the air quality colour scale
and the preparedness hub's save-and-share logic. The whole suite runs in a few
seconds and passes cleanly.

% ============================ 7. ACCURACY & LIMITS ============================
\section{Accuracy, Honesty, and Limits}

A core principle of the project was to be clear about what the data and models
can and cannot do. The main points a reader should know:

\begin{itemize}[leftmargin=1.4em, itemsep=0.3em]
\item \textbf{The risk map is a planning aid, not a warning.} Differences
between map cells come from each area's fire history, not from separate local
weather. The official fire-danger rating is shown next to it for comparison.
\item \textbf{The climate projections are a placeholder.} They have the right
shape for the page but are not the live download yet. Swapping in the real data
is a single file replacement with no code change, and the page says so.
\item \textbf{Air quality forecasting covers one location.} It cannot see smoke
arriving from outside the region until local readings start to rise.
\item \textbf{Historical fire records and the risk map cover the
Thompson-Okanagan.} The live hazard layers (fires, heat detections, evacuation,
fire weather, air quality, smoke) cover the whole province.
\end{itemize}

\callout{The Fire Weather Index used here is computed with Canada's official
equations. The air quality health figures follow Health Canada's formula. Where
the platform calculates something itself, it uses the recognised standard.}

% ============================ 8. STATUS ============================
\section{Status and Summary}

The four features and both models are built, connected to live data, and
working. The platform passes its full test suite (47 back-end and 22 front-end
checks) with a clean build, and the code, documentation, and model cards are
organised for handoff.

The table below maps each proposal commitment to its current state.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}L{7.5cm}L{5.5cm}@{}}
\toprule
\textbf{Proposal commitment} & \textbf{Status} \\
\midrule
Wildfire Risk Map (daily zones + current and past fires) & Built \\
Air Quality Monitor (live reading, 48-hour forecast, smoke calendar) & Built \\
Community Preparedness Hub (checklist, evacuation status, points) & Built \\
Climate Trend module (history + projections, optional TRU carbon) & Built \\
Two models validated on held-out 2022 and 2023 & Done \\
Free and public data only, no personal data stored & Met \\
Open-source release & Done (MIT) \\
Final research report & This document \\
Public online hosting & Ready to deploy; not yet hosted \\
\bottomrule
\end{tabular}
\caption{Proposal commitments and their status.}
\end{table}

A few items remain before a public launch: hosting the platform online,
recording a short demo video, on-device tablet testing, and swapping the
placeholder climate projections for the live download. The climate history
covers the full record since 1999 (27 years); the round ``30-year'' figure in
the proposal is reached as more recent seasons are added each year. Two
proposal ideas were left for future work: adding satellite vegetation health to
the risk model, and the economic study of smoke and healthcare demand. Neither
blocks use of the platform today.

WildfireIQ Kamloops shows that a single person, using only free and open data,
can build a clear and honest wildfire information tool for their community. The
result is a platform a resident can check on a smoky afternoon, and one a
researcher or official could cite.

\vspace{1em}
{\noindent\color{ember}\rule{\linewidth}{1pt}}
\vspace{0.4em}

\noindent{\small\color{slate} Project repository and full documentation are
included with this report. See the README and the \texttt{documents} folder for
the engineering log, data dictionary, and model cards.}

\end{document}
